Influence of journal data archiving mandates on author attitudes, experiences,

### 2018-11-21 manuscript.Rmd
make more past tense, add start of abstract, fix references

---
title: Influence of journal data archiving mandates on author attitudes, experiences,
  and practices
author: "Heather A. Piwowar, Todd J. Vision"
csl: peerj.csl
output:
  md_document:
    variant: markdown_github
  html_document:
    css: style.css
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  word_document:
    fig_caption: yes
bibliography: input_references.bib
---

@@

@@

Sharing data is a tenet of science, yet commonplace in only a few subdisciplines.  Publicly available datasets have many potential reuses.  Unfortunately, these advantages only indirectly benefit the stakeholders who bear most of the costs for sharing their datasets: the primary data-producing investigators. As a result, authors often actively or passively withhold their research datasets from other investigators  
`r citep(c(bib["blumenthal1997withhol"], bib["ochsner2008much-ro"], bib["savage2009empiric"]))`.

@@

A journal policy that requires authors to share the datasets that support their results as a condition of publication may impact the practices, attitudes, and opinions of data producing authors in many ways. The most direct impact is likely a change in frequency of data sharing, and a shift to data archiving locations specified in the policy.  Early studies suggest that authors who publish in journals with data sharing policies are more likely to share their data than authors who publish similar studies in similar journals `r citep(c(bib["piwowar2008a-revie"], bib["Piwowar_2011"]))`, but these investigations were correlative and did not attempt to directly measure the impact of journal policies over time.  Previous work emphasizes the importance of sharing data through best-practice archiving methods and locations.  Authors who share data upon request often discriminate against use and users `r citep(c(bib["campbell2008what-ar"], bib["reidpath2001data-sh"]))`.  The permanence of data sharing mechanisms is also a concern; email and lab-website based sharing choices have been shown to be impermanent solutions at best `r citep(c(bib["wren2008url-dec"], bib["wren2006e-mail"]))`.

@@

A few previous studies have estimated the impact of sharing data on the experiences of authors `r citep(c(bib["blumenthal2006data-wi"], bib["gleditsch2003posting"], bib["piwowar2007sharing"], bib["ventura2005mandato"]))`, but there is much to be learned in this area to ground expectations, accurately estimate costs and benefits, and learn about ways policies could be improved to enhance positive experiences and avoid negative ones.

@@

## Upcoming policy change is a useful opportunity for policy evaluation

Several journals plan to adopt the Joint Data Archiving Policy in early 2011, as described at http://datadryad.org/jdap.

This is a great opportunity to study the effect of journal data sharing mandates on the attitudes, behaviour, experiences of authors to help inform future policy decisions.

The JDAP policy requires data sharing in a public archive. Early evidence suggests stronger policies are associated with a higher frequency of data sharing `r citep(c(bib["piwowar2008a-revie"], bib["Piwowar_2011"]))`.  Since JDAP requires data sharing as a condition of publication, studying this policy adoption is an opportunity to possibly measure the maximum likely effect.  Also, since the adopting journals include almost all of the high-impact journals in the field of Evolutionary Science (as per the 2009 ISI JCR), it is likely that even authors who dislike the policies will continue publishing in these journals and possibly participate in our measurement of opinions.

@@

Because measurements of actual dataset archiving behavior can be accurately measured with a retrospective analysis of research artifacts, our proposed survey does not emphasize data archiving behaviour.  Instead, the questionnaire will focus on information that is difficult or impossible to collect without a timely survey of data-producing investigators.

@@

Although some data sharing mandates appear to have achieved almost universal sharing `r citep(bib["noor2006data-sh"])`, this is not always the case `r citep(bib["ochsner2008much-ro"])`.  Several surveys have asked about author’s perceived obstacle to sharing data `r citep(c(bib["hedstrom2008researc"], bib["scherle2008buildin"]))`, but changes in these attitudes have not been monitored across a policy change.  To our knowledge, the impact of data sharing mandates have only been subject to one evaluation:  the editors of Physiological Genomics briefly surveyed their authors and reviewers two years after instituting a policy that required public archiving of gene expression microarray data `r citep(bib["ventura2005mandato"])`, though unfortunately no baseline measurements are available about attitudes before the policy.  Interestingly, a recent large-scale evaluation `r citep(bib["Piwowar_2011"])` into the prevalence of microarray data sharing found this same journal to have the highest rate of public archiving.

@@

The instrument used to collect data is be an online questionnaire. The questionnaire is designed to elicit the respondent’s practices, experiences, and opinions around public archiving of research datasets by asking about their opinion of community norms, recent behaviour with respect to a recently published paper, previous experiences with data sharing, and attitudes toward hypothetical journal policies

## Control jounrnals and sample size

The proposed study is subject to two different types of uncertainty:  uncertainty about the estimation of aggregate parameters, and also “uncertainty about the ability of the control group to reproduce the counterfactual of how the treated unit would have evolved in the absence of the treatment” (`r citep(bib["abadie2010synthet"])`. It was difficult to know how to conduct a formal power analysis in these conditions, so we  based our sample size decisions on the following information.

For populating the “treatment arm,” we asked all corresponding authors of all JDAP-adopting journals.

For populating the “control arms,” we recognized that the uncertainty is predominantly determined by the applicability of the controls.  Little information is currently available to determine, a priori, which journals will serve as appropriate controls.  As such, we decided that surveying twice as many control journals as treatment journals was an appropriate rule of thumb.  This balanced the need to avoid overpowering the study (and thereby increasing survey-fatigue and effort) with a desire to ensure the most useful results (anticipating that unexpected future events may render some controls inappropriate during the three years of this study, as well as recognizing that the quality of inference improves as the number of available comparison units increases `r citep(bib["abadie2010synthet"])`.

Control journals were selected by manual selection from a set of journals identified through ISI Journal Citation report categories as well as co-citation lists.  Attempts were made to choose journals with comparable impact factors and inclination to produce quantitative datasets.

Assuming a 20% response rate and a control arm that was twice as large as the treatment arm, we expected a maximum of 165 responses per month, or almost 6000 responses over the course of the three year study. The actual number of responses was expected to be lower than this because only invited each corresponding author the first time they published a paper.

@@

The email invitation included  a URL where subjects can learn more about the study and then read a fact sheet, consent, and participate if they choose.  This email contained contain an “opt out” option, which will remove their address from the reminder list.  Those who did not “opt out” were contacted by email again, a week later, politely asking them to take the survey if they haven’t already.

@@

The primary analysis will be based on responses to be a seven-point Likert scale asking how strongly authors believe:  “It is the community norm in my field to publicly share datasets upon study publication by archiving all datasets online, for use by anyone for any purpose.”  Our goal is to see if author's responses to this question were influenced by publishing in a journal that had implemented JDAP.

@@

Journal policies do help  `r citep(c(bib["vines2013mandate"], bib["Piwowar_2011"], bib["alsheikh-ali2011public"]))`, though often not sufficient to achieve data availability for the majority of studies `r citep(c(bib["Piwowar_2011"], bib["alsheikh-ali2011public"], bib["roche2015public"], bib["stodden2018an-empi"], bib["naudet2018data-sh"]))`.

@@

Our findings are consistent with previous studies that found authors were more likely to share data if they had prior experience sharing or reusing data: `r citep(c(bib["Piwowar_2011"]))`.

@@

@@

# Acknowlegements

Everyone who took the survey.

@@

Heather's postdoc funding.

>>>>

---
title: Influence of journal data archiving mandates on author attitudes, experiences,
  and practices
author: "Heather A. Piwowar, Todd J. Vision"
csl: peerj.csl
output:
  word_document:
    fig_caption: yes
  md_document:
    variant: markdown_github
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  html_document:
    css: style.css
    fig_caption: yes
    keep_md: yes
    number_sections: yes
    toc: yes
bibliography: input_references.bib
---

@@

Several journals adopted the Joint Data Archiving Policy in early 2011, as described at http://datadryad.org/jdap.

This was a great opportunity to study the effect of journal data sharing mandates on the attitudes, behaviour, experiences of authors to help inform future policy decisions.

The invitations were sent out monthly between February 2011 and March 2014 to a total of 23853 23853  authors, and 4102 4102 responded (corresponding to a response rate of 17%).

the introduction of the JDAP data archiving policy significantly increases the odds of a positive response regarding data sharing by a factor of 1.36
1.36
 (CI=[1.16;1.60]
1.16
1.60
), while publishing date and whether the journal ever adopted a JDAP policy did not show any statistically significant effects.

We found the similar results when we looked at the author’s responses to community norms about public data archiving rather than peer-to-peer data sharing

this study only facilitates a relatively short-term look at policy effect: it will likely take many more years for the effect of these policies to be fully felt.

@@

Sharing data is a tenet of science, yet commonplace in only a few subdisciplines.  Publicly available datasets have many potential reuses.  Unfortunately, these advantages only indirectly benefit the stakeholders who bear most of the costs for sharing their datasets: the primary data-producing investigators. As a result, authors often actively or passively withhold their research datasets from other investigators  
`r citep(c(bib["blumenthal1997withhol"], bib["ochsner2008much"], bib["savage2009empiric"]))`.

@@

A journal policy that requires authors to share the datasets that support their results as a condition of publication may impact the practices, attitudes, and opinions of data producing authors in many ways. The most direct impact is likely a change in frequency of data sharing, and a shift to data archiving locations specified in the policy.  Early studies suggest that authors who publish in journals with data sharing policies are more likely to share their data than authors who publish similar studies in similar journals `r citep(c(bib["piwowar2008a-revie"], bib["Piwowar2011"]))`, but these investigations were correlative and did not attempt to directly measure the impact of journal policies over time.  Previous work emphasizes the importance of sharing data through best-practice archiving methods and locations.  Authors who share data upon request often discriminate against use and users `r citep(c(bib["campbell2008what-ar"], bib["reidpath2001data-sh"]))`.  The permanence of data sharing mechanisms is also a concern; email and lab-website based sharing choices have been shown to be impermanent solutions at best `r citep(c(bib["wren2008url-dec"], bib["wren2006e-mail"]))`.

@@

A few previous studies have estimated the impact of sharing data on the experiences of authors `r citep(c(bib["blumenthal2006data-wi"], bib["gleditsch2003posting"], bib["piwowar2007sharing"], bib["ventura2005"]))`, but there is much to be learned in this area to ground expectations, accurately estimate costs and benefits, and learn about ways policies could be improved to enhance positive experiences and avoid negative ones.

@@

## Policy change is a useful opportunity for policy evaluation

Several journals adopted the Joint Data Archiving Policy in early 2011, as described at http://datadryad.org/jdap.

This was a great opportunity to study the effect of journal data sharing mandates on the attitudes, behaviour, experiences of authors to help inform future policy decisions.

The JDAP policy requires data sharing in a public archive. Early evidence suggests stronger policies are associated with a higher frequency of data sharing `r citep(c(bib["piwowar2008a-revie"], bib["Piwowar2011"]))`.  Since JDAP requires data sharing as a condition of publication, studying this policy adoption is an opportunity to possibly measure the maximum likely effect.  Also, since the adopting journals include almost all of the high-impact journals in the field of Evolutionary Science (as per the 2009 ISI JCR), it is likely that even authors who dislike the policies will continue publishing in these journals and possibly participate in our measurement of opinions.

@@

Because measurements of actual dataset archiving behavior can be accurately measured with a retrospective analysis of research artifacts, our proposed survey does not emphasize data archiving behaviour.  Instead, the questionnaire focused on information that is difficult or impossible to collect without a timely survey of data-producing investigators.

@@

Although some data sharing mandates appear to have achieved almost universal sharing `r citep(bib["noor2006data-sh"])`, this is not always the case `r citep(c(bib["ochsner2008much"]))`.  Several surveys have asked about author’s perceived obstacle to sharing data `r citep(c(bib["hedstrom2008researc"]))`  `r citep(c(bib["scherle2008buildin"]))`, but changes in these attitudes have not been monitored across a policy change.  To our knowledge, the impact of data sharing mandates have only been subject to one evaluation:  the editors of Physiological Genomics briefly surveyed their authors and reviewers two years after instituting a policy that required public archiving of gene expression microarray data `r citep(c(bib["ventura2005"]))`, though unfortunately no baseline measurements are available about attitudes before the policy.  Interestingly, a recent large-scale evaluation `r citep(c(bib["Piwowar2011"]))` into the prevalence of microarray data sharing found this same journal to have the highest rate of public archiving.

@@

The instrument used to collect data was an online questionnaire. The questionnaire was designed to elicit the respondent’s practices, experiences, and opinions around public archiving of research datasets by asking about their opinion of community norms, recent behaviour with respect to a recently published paper, previous experiences with data sharing, and attitudes toward hypothetical journal policies.

## Control journals and sample size

The study was subject to two different types of uncertainty:  uncertainty about the estimation of aggregate parameters, and also “uncertainty about the ability of the control group to reproduce the counterfactual of how the treated unit would have evolved in the absence of the treatment” (`r citep(bib["abadie2010synthet"])`.

It was difficult to know how to conduct a formal power analysis in these conditions, so we  based our sample size decisions on the following information:

- For populating the “treatment arm,” we asked all corresponding authors of all JDAP-adopting journals.

- For populating the “control arms,” we recognized that the uncertainty is predominantly determined by the applicability of the controls.  Little information is currently available to determine, a priori, which journals will serve as appropriate controls.  As such, we decided that surveying twice as many control journals as treatment journals was an appropriate rule of thumb.  This balanced the need to avoid overpowering the study (and thereby increasing survey-fatigue and effort) with a desire to ensure the most useful results (anticipating that unexpected future events may render some controls inappropriate during the three years of this study, as well as recognizing that the quality of inference improves as the number of available comparison units increases `r citep(bib["abadie2010synthet"])`.

- Control journals were selected by manual selection from a set of journals identified through ISI Journal Citation report categories as well as co-citation lists.  Attempts were made to choose journals with comparable impact factors and inclination to produce quantitative datasets.

- Assuming a 20% response rate and a control arm that was twice as large as the treatment arm, we expected a maximum of 165 responses per month, or almost 6000 responses over the course of the three year study. The actual number of responses was expected to be lower than this because only invited each corresponding author the first time they published a paper.

@@

The email invitation included a URL where subjects could learn more about the study and then read a fact sheet, consent, and participate if they choose.  This email contained contain an “opt out” option, which will remove their address from the reminder list.  Those who did not “opt out” were contacted by email again, a week later, politely asking them to take the survey if they haven’t already.

@@

The primary analysis was based on responses on a seven-point Likert scale asking how strongly authors believe:  “It is the community norm in my field to publicly share datasets upon study publication by archiving all datasets online, for use by anyone for any purpose.”  Our goal was to see if author's responses to this question were influenced by publishing in a journal that had implemented JDAP.

@@

Journal policies do help  `r citep(c(bib["vines2013mandate"], bib["Piwowar2011"], bib["alsheikh-ali2011public"]))`, though often not sufficient to achieve data availability for the majority of studies `r citep(c(bib["Piwowar2011"], bib["alsheikh-ali2011public"], bib["roche2015public"], bib["stodden2018an-empi"], bib["naudet2018data-sh"]))`.

@@

Our findings are consistent with previous studies that found authors were more likely to share data if they had prior experience sharing or reusing data: `r citep(c(bib["Piwowar2011"]))`.

@@

Also work in these:
- The influence of journal submission guidelines on authors' reporting of statistics and use of open research practices: `r citep(c(bib["giofre2017the-inf"]))`
- Journal Data Sharing Policies and Statistical Reporting Inconsistencies in Psychology: `r citep(c(bib["nuijten"]))`
- 10.1371/journal.pone.0194768

@@

# Acknowlegements and Notes

We'd like to sincerely thank everyone who took the survey.

The first paragraphs of the introduction are verbatim from Piwowar, Day & Fridsma (2007); its original publication was under a CC-BY license.

The authors finally completed the analysis and posted this preprint more than five years after the study (and Heather's postdoc) ended: better late than never.  The introduction has been only lightly revised since it was written 5 years ago; regretfully, important recent studies have no doubt been overlooked.

@@

This study was funded by DataONE (OCI-0830944) and Dryad (DBI-0743720). The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
